working owls methods

### 2020-11-11 manuscript_working/ms_working_RMD.Rmd
updating manuscript with richness results/methods

--- 
title: "working owls methods"
author: "Althea Archer"
date: "11/10/2020"
output:
  word_document:
    fig_caption: yes
  pdf_document:
    includes:
      latex_engine: pdflatex
    keep_tex: yes
csl: ecology.csl
bibliography: owl_references.bib
---

@@

@@

\[
\begin{aligned}
\text{E}[\psi_{t,h}] &= \mu^\psi_{t,h} = a^\psi_{t,h}/(a^\psi_{t,h}+b^\psi_{t,h})\\
\rho^\psi_{t,h} &= a^\psi_{t,h}+b^\psi_{t,h}\\
\mu^\psi_{t,h} &\sim \text{Uniform}(0.01, 0.99)\\
\rho^\psi_{t,h} &\sim \text{N}(5,1)\text{, truncated to 0.01,10}\\
\end{aligned}
\]

We assumed that detecting an owl depended on an owl being present during that survey ($z_{t,h,i} = 1$) and the probability of detection, which was related to the broadcast species:

@@

where generally $\text{logit}(p_{t,h,i,j,k}) = \beta_kX_k$, or more specifically:

@@

This model provides a consistent probability of detection for all surveys in the first two minutes of each survey before the broadcast owl recording was played. Then, the probability of detection for each post-broadcast time period would depend on the species of broadcast owl. This allowed species-specific behavior in response to the different broadcast owl species [@Baumgardt:2019]. We used means parameterization such that the coefficients were interpretable as the effect of that specific broadcast species $k$ (including the pre-broadcast time frame). The priors for every logistic model coefficient $\beta_k$ were chosen to be near uniform as recommended in @Gelman:2008. Specifically, we used the Cauchy prior as such: $\beta_k \sim \text{Cauchy}(\text{precision} = 0.16)$.

@@

We used the R2jags package in R [@R:2014,@Plummer:2013] to implement the occupancy model for three owl species: Mottled, Spectacled, and FerPy owls. These owls had enough positive detections to analyze occupancy, as Mottled, Spectacled, and FerPy owls had 542, 137, and 187 positive detections over the 11 year period, respectively. Based on our understanding of owl ecology, we assumed that Spectacled and FerPy owls would not occupy route M1 in Montecristo, so we removed that route from those occupancy models.

@@

To model species richness at each route in each year, we assumed that there were 14 possible species present in El Salvador. This included the 10 species observed during our surveys (including unknown owl species) and an additional 4 unidentified species. This limit to species richness relates directly to the 13 species known to inhabit El Salvador (cite Owls of El Salvador here) plus one species we identified (Stygian owl), which was previously undocumented in El Salvador. We augmented our owl detection data with 4 additional potential species with all-zero detection [@Royle:2007]. For each species $s$ (where $s = 1,2, \ldots,14$), $w_{t,h,s}$ is a binary indicator of that owl species being present in each route and year:

$$
w_{t,h,s} \sim \text{Bernoulli}(\Omega_h)
$$
where $\Omega_h$ is the probability of species presence in each route $h$. Now, the presence of an owl species in any route, year, and survey, $z_{t,h,i,s}$, is conditional on the probability of occupancy for that route, year, and species $\psi_{t,h,s}$ *and* that species belong to that route's community ($w_{t,h,s}$):

@@

Richness is conditional on each species presence and was estimated as the sum of owl species present in each route and year:

@@

\[
\begin{aligned}
\psi_{t,h,s} &\sim \text{Beta}(a^\psi_{t,h,s},b^\psi_{t,h,s})\\ 
\text{logit}(\mu^\psi_{t,h}) &= \text{E}(\psi_{t,h,s}) \sim N(0,2.25)\\
\ln(\rho^\psi) &\sim N(5,1)\text{, truncated to (0.01,10)}\\
a^\psi_{t,h,s} &= \mu^\psi_{t,h}\rho^\psi\\
b^\psi_{t,h,s} &= \rho^\psi - \mu^\psi_{t,h}\rho^\psi\\
\end{aligned}
\]

@@

Consistent with our approach for the occupancy models, we assumed that detecting an owl depended on an owl of that species being present during that survey ($z_{t,h,i,s} = 1$) and the probability of detection $p_{t,h,i,j,k,s})$, which was related to the broadcast species:

@@

We used the R2jags package in R [@R:2014,@Plummer:2013] to implement the richness model. 
We ran 3 chains for 20,000 iterations and 2000 iterations discarded as burn-in and a thinning rate of 2, for a total of 27,000 iterations comprising the posterior distributions for each model parameter. Initial values for $z_{t,h,i,s}$ were equivalent to one if that species of owl $s$ was detected in that survey $i$ and otherwise zero. Similarly, initial values for $w_{t,h,s}$ were equivalent to one if that species of owl was detected in that route and year. We visually inspected traceplots to verify that chains mixed well (Supplemental information S4xx).

@@

>>>>

--- 
title: "working owls methods"
author: "Althea Archer"
date: "11/11/2020"
output:
  word_document:
    fig_caption: yes
  pdf_document:
    includes:
      latex_engine: pdflatex
    keep_tex: yes
csl: ecology.csl
bibliography: owl_references.bib
---

@@

``` {r}
load("../data/output_data/richness_species_accounts_byRt.Rdata")
library(pander)
pander(species.accounts.byRt)
```

@@

\[
\begin{aligned}
\text{E}[\psi_{t,h}] &= \mu^\psi_{t,h} = a^\psi_{t,h}/(a^\psi_{t,h}+b^\psi_{t,h})\\
\rho^\psi_{t,h} &= a^\psi_{t,h}+b^\psi_{t,h}\\
\text{logit}(\mu^\psi_{t,h}) &\sim \text{N}(0, 2.25)\\
\ln(\rho^\psi_{t,h}) &\sim \text{N}(5,1)\text{, truncated to 0.01,10}\\
\end{aligned}
\]

We assumed that detecting an owl depended on an owl being present during that survey ($z_{t,h,i} = 1$) and the probability of detection, which was related to the species of owl played for broadcast ($k$):

@@

where, generally, $\text{logit}(p_{t,h,i,j,k}) = \beta_kX_k$, or more specifically:

@@

This model provides a consistent probability of detection for all surveys in the first two minutes of each survey before the broadcast owl recording was played ($k = \text{Pre-broadcast}$). Then, the probability of detection for each post-broadcast time period would depend on the species of broadcast owl. This allowed species-specific behavior in response to the different broadcast owl species [@Baumgardt:2019]. We used means parameterization such that the coefficients were interpretable as the effect of that specific broadcast species $k$ (including the pre-broadcast time frame). The priors for every logistic model coefficient $\beta_k$ were chosen to be near uniform as recommended in @Gelman:2008. Specifically, we used the Cauchy prior as such: $\beta_k \sim \text{Cauchy}(\text{precision} = 0.16)$.

@@

We used the R2jags package in R [@R:2014,@Plummer:2013] to implement the occupancy model for three owl species: Mottled, Spectacled, and FerPy owls. These owls had enough positive detections to analyze occupancy, as Mottled, Spectacled, and FerPy owls had 542, 137, and 187 positive detections over the 11 year period, respectively. Based on our understanding of owl ecology, we assumed that Spectacled and FerPy owls would not occupy route M1 in Montecristo, so we removed M1 from those species' occupancy models.

@@

To model species richness at each route in each year, we assumed that there were 14 possible species present in El Salvador. This included the 10 species observed during our surveys (including unknown owl species) and an additional 4 unidentified species. This limit to species richness relates directly to the 13 species known to inhabit El Salvador (cite Owls of El Salvador here) plus one species we identified (Stygian owl), which was previously undocumented in El Salvador. We augmented our owl detection data with 4 additional potential species with all-zero detection [@Royle:2007]. For each species $s$ (where $s = 1,2, \ldots,14$), $w_{t,h,s}$ was a binary indicator of that owl species being present in each route and year:

$$
w_{t,h,s} \sim \text{Bernoulli}(\Omega_h)
$$
where $\Omega_h$ was the probability of species presence in each route $h$. Now, the presence of an owl species in any route, year, and survey, $z_{t,h,i,s}$, was conditional on the probability of occupancy for that route, year, and species $\psi_{t,h,s}$ *and* that species belong to that route's community ($w_{t,h,s}$):

@@

Richness was conditional on each species presence and was estimated as the sum of owl species present in each route and year:

@@

\[
\begin{aligned}
\psi_{t,h,s} &\sim \text{Beta}(a^\psi_{t,h,s},b^\psi_{t,h,s})\\ 
a^\psi_{t,h,s} &= \mu^\psi_{t,h}\rho^\psi\\
b^\psi_{t,h,s} &= \rho^\psi - \mu^\psi_{t,h}\rho^\psi\\
\text{logit}(\mu^\psi_{t,h}) &= \text{E}(\psi_{t,h,s}) \sim N(0,2.25)\\
\ln(\rho^\psi) &\sim N(5,1)\text{, truncated to (0.01,10)}\\
\end{aligned}
\]

@@

Consistent with our approach for the occupancy models, we assumed that detecting an owl depended on an owl of that species being present during that survey ($z_{t,h,i,s} = 1$) and the probability of detection $p_{t,h,i,j,k,s}$, which was related to the broadcast species:

@@

We used the R2jags package in R [@R:2014,@Plummer:2013] to implement the richness model. 
We ran 3 chains for 20,000 iterations and 2000 iterations discarded as burn-in and a thinning rate of 2, for a total of 27,000 iterations comprising the posterior distributions for each model parameter. Initial values for $z_{t,h,i,s}$ were equivalent to one if that species of owl $s$ was detected in that survey $i$ and otherwise zero. Similarly, initial values for $w_{t,h,s}$ were equivalent to one if that species of owl was detected in that route and year and otherwise zero. We visually inspected traceplots to verify that chains mixed well (Supplemental information S4xx).

@@

## Richness Model Results

## Detection Results
